Improving the communication and accessibility of stock assessment using interactive

### 2019-05-06 analysis/paper/manuscript.Rmd
Another round of edits

---
title: Improving the communication and accessibility of stock assessment using interactive
  visualization tools
author: "Paul M. Regular, Gregory J. Robertson, Keith P. Lewis, etc?"
date: "Fisheries and Oceans Canada, Northwest Atlantic Fisheries Center, 80 East White Hills, St. John's, Newfoundland and Labrador, A1C 5X1, Canada"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
bibliography: references.bib
---

Quantitative stock assessment plays a central role in modern fisheries management. Over time and as new methods are developed, biologists have more and more data available to inform stock assessments. These data might be richer information on stock structure using an increasing array of markers and biomarkers, and/or the continued lengthening of population status and catch time series. Concurrently, analysis methods have evolved with techniques now available that integrate many data sources into one stock assessment model. For well-monitored stocks, the challenge has shifted from having sufficient data and information to provide sound advice on stock status, to presenting the mass of information produced by increasingly complex statistical models to end-users in a meaningful way. Traditional methods of presenting information at stakeholder meetings and to fisheries managers are simply no longer sufficient to convey the richness of the information. More importantly, static presentation formats tend to stifle meaningful discussion of model results, and important questions and assumptions of the models that should be discussed tend to get lost under the sheer volume of output. A solution to this problem of communicating complex output and results to encourage engaging discussion can be through using the effective use of interactive visualization tools. These are common tools we use every day on a range of web sites, but their use is no longer restricted to web site developers as these tools are being integrated into software commonly used by the research community.

Here we aim to demonstrate how interactive visualization tools provide an efficient and effective means of exploring and communicating the ever expanding array of data inputs and model outputs. First, we focus on data that are commonly used in stock assessment and provide one example where interactive maps were used to simplify the detailed exploration of data from a long-term tagging program. Second, we focus on the modeling aspect of stock assessment and, again, we use an example to demonstrate how dynamic and interactive data visualizations can be used to explore, diagnose and communicate results from an integrated assessment model. This structure corresponds to the two general steps in the stock assessment process: 1) knowing the data, and 2) data modeling.

@@

The data sets used in stock assessments are constantly growing. This growth in data either stems from the continuation of long-term monitoring efforts or from the addition of new monitoring programs. As such, stock assessment biologists often have to manage large volumes of data from a variety of sources. Time series of reported landings and catch-at-age are often analyzed in conjunction with data from “fishery-independent” surveys that track changes in abundance and, in many cases, also monitor trends in biological factors such as age composition, growth rates, sex ratios and maturation stages. For some data-rich stocks, mark and recapture studies are also carried out to estimate movement, migration, growth rate, natural mortality, and discard mortality. All of the above-mentioned data sets are multidimensional and as the volume and variety of these data increases, it becomes more difficult to be aware of the details of each data set and discover key patterns within each. This challenge can be mediated, to a degree, by the application of interactive visualization tools as that they allow detailed exploration of the data behind the plot. For instance, the ability to zoom in on features or areas of interest and hover over specific points to reveal more information creates an interactive user-driven experience that expedites explorations of the data. This is exemplified by an interactive mapping tool developed for the exploration of a long-term tagging study.

Northern cod (NAFO Divisions 2J3KL) has a rich history of tagging, starting in in 1954 [@taggart1995] and continuing to this day. The tagging and recovery data is captured in standardized database, with fields typical of most tagging programs. This data base has over 600,000 records as of early 2019, with 2,000-10,000 tags deployed annually in recent decades. The tagging and capture data are used in the current assessment model for this stock [@cadigan2016], but tools to explore this extensive data set were limited, especially from a spatial perspective. To begin to explore and understand this large data set, we built a simple shiny [@chang2018a] application using the mapping package leaflet [@cheng2018], all within the framework of the R programming language [@R] and the RStudio IDE [@Rstudio]. As the application developed, it was incorporated into a shinydashboard [@chang2018b], as a means to quickly and dynamically subset the data (e.g. ranges of release and capture years, specific geographic locations). Given the large number of tags released, often at nearby sites, visualizing the data with static mapping was particularly challenging. The markercluster (https://github.com/Leaflet/Leaflet.markercluster) function available in leaflet was particularly useful as a means to dynamically scale the level of pooling of spatial points (Figure \@ref(fig:map)). This basic mapping tool allowed us to quickly become familiar with the data, identify outliers and incorrect data entries, and explore options on how to spatially pool the data for subsequent demographic analysis. Further tabs were added to provide basic summaries of the selected data (Figure \@ref(fig:summary)).

```{r map, echo = FALSE, fig.cap = '- Screen shot of cod tag mapping tool using shinydashboard and leaflet. The markerCluster function dynamically splits or pools tagging locations (red, orange, green or yellow points) depending on zoom level, the recoveries positions (blue) are much fewer, and are left to be plotted individually at all scales. Options to include pop up labels are included, so specific information on each point can be retrieved with a mouse click (in this case: tag number, fish length, date released and date captured), which is particular useful when error checking.'}
knitr::include_graphics("figures/tagging_dashboard_map_tab.png")
```

@@

The Northern cod stock off southern Labrador and eastern Newfoundland is one of the most well studied stocks in eastern Canada, perhaps by virtue of its history. As such, there are multiple monitoring programs that help inform the status of the stock and data from most of these programs have been integrated into a state-space stock assessment model, called NCAM [@cadigan2016]. The "base case" model includes information from research vessel autumn trawl surveys (1983-present), Sentinel fishery surveys (1995-present), inshore acoustic surveys (1995-2009), fishery catch-at-age compositions and partial fishery landings (1983-present), and tagging (1983-present). Using a series of observation equations, this TMB based model reduces thousands of historical data points from these monitoring programs into quantities such as recruitment, spawning stock biomass, fishing mortality and natural mortality. Once the model is fit to the data, the next step is to produce visual representations of the data and model. A standard approach would be to produce static presentations and documents with a series of figures and tables. With large amounts of model inputs and outputs, this approach quickly becomes overwhelming for both the analyst and all the stakeholders involved in the stock assessment process. First, it is not feasible for the analyst to include and describe all figures and tables produced in a single document. Second, it is difficult for stakeholders to efficiently digest the information that has been compressed into a series of static slides or pages. Interactive documents provide a potential solution to this problem as they allow much more information to be contained and accessible on a single screen.

In the pursuit for an easier and more efficient way to communicate results from NCAM, an interactive and self-contained "dashboard" was developed for the 2018 assessment of Northern cod [@dwyer2019]. The concept of using a dashboard was burrowed from the business community where dashboards are frequently used to group a series of interactive visuals and tables to provide at-a-glance views of key performance indicators. We used R-based packages [@R] and the RStudio IDE [@Rstudio] to construct a tool for exploring the input and output of NCAM, specifically the flexdashboard [@allaire2017] package to group interactive plotly-based [@sievert2018] visuals into a dynamic document. We also used the crosstalk [@cheng2016] package to link the data displayed across multiple plots. Via R Markdown [@allaire2018], the dashboard is rendered into a self-contained html file that is reproducible, interactive, and easy to update following modifications to the model or the addition of new data.

@@

```{r rv, echo = FALSE, fig.cap = '- Screenshot of the "RV survey" tab from the NCAM dashboard where total observed (dots) and NCAM model predicted values (lines) for the DFO RV survey index are shown in the upper left panel and scaled matrix plot of age-disaggregated standardized log residuals are shown in the lower left panel (blue = positive, red = negative, symbols scaled by size; grey = index values of zero). Age-disaggregated observed (dots) and predicted (lines) values from the DFO RV survey are shown in the right panel. The tab on the right panel including residual plots are hidden here, but assessable in the dashboard.'}
knitr::include_graphics("figures/NCAM_dashboard_rv-survey_tab.png")
```

@@

The amount of data available to scientists has grown by orders of magnitude in recent decades as has the complexities of data management, exploratory data analysis, as well as the formal analyses and associated diagnostics. The majority of this sequence of events, sometimes called “the data pipeline”, have not traditionally been part of the peer-review process which sees only the end products. However, decisions made at these steps increasingly influence the outcome of the study. Gelman and Loken [-@gelman2014] coined the term "garden of forking paths" to illustrate that different conclusions can be arrived at depending on what decisions one makes along the garden path. Due to a number of limitations, such as available pages in journals, much of the data pipeline is not transparent nor is it reproducible, even if it is one’s own analysis. A number of authors have recently advocated for a culture of open science and reproducible research, i.e., a change in the transparency and reproducibility of science. Proponents of open science and reproducible research posit a number of benefits including a more productive and responsible scientific culture, an ability to address larger and more complex questions, as well as a more efficient workflow and ability to reproduce one’s own work [@lewis2018].

We submit that the dashboard tool is also a step forward in terms of open science and reproducible research in addition to being a major step forward in how stock assessments are presented and critiqued. Consider the following process. Through the Canadian Science Advisory Secretariat (CSAS), DFO Science provides peer-reviewed science advice to fisheries managers at regular meetings (Regional Assessment Process). These meetings are attended by a variety of stakeholders including harvesters, academics, and NGOs. All participants attending a CSAS meeting are encouraged to question, comment and constructively challenge the science presented and the goal is to provide the best possible science advice to the Minister, managers, stakeholders and the public. While the entire data-pipeline can be revealed to all by sharing code and data, it is unlikely that an individual has the time to fully review and recreate a particular analysis in its entirety when most participants are already overcommitted [@banks2011]. In the case of NCAM, considerable statistical experience and expertise is required simply to run the model (or perhaps to even tweak the model). However, the dashboard removes many of these difficulties. The html file contains all model inputs, outputs and diagnostics in an open and easy to use interactive document. The number of analyses can be expanded upon request and can be distributed prior to meetings and participants can assess the results and diagnostics at their leisure. Specific experience with the tools used to generate the results are not required to constructively and critically review of the results presented in the dashboard.

We must acknowledge there are costs to adopting this approach of using the interactive tools. The first and obvious one being that staff are required to have both the time and training to effectively to develop these tools. In our two examples, the developers were relatively proficient with R and have natural aptitudes for programming, but had their formal training in field-based population ecology. Most programming issues were readily addressed through simple internet searchers, but some learning time was certainly needed (a few weeks). As more people in the fisheries community develop and use these tools, we expect tool develop speed and quality will only increase. It is our experience that the upfront investment in time has already paid dividends in terms of efficiencies of delivering products for subsequent stock assessments, and in the quality of the data exploration and understanding. Scientific exploration is rarely a linear process, and the ability to jump around a dashboard is a more in tune with the natural way scientist’s pursue ideas. In contrast, a slide presentation is effectively a linear and a single way of developing a scientific narrative; a method hardly conducive to exploring complex data problems.

In summary, we submit that the dashboard tool is an important step forward in transparent and reproducible science. Further, this tool makes stock assessments more accessible to stakeholders. Beyond stock assessments, interactive documents could be used for a more fulsome review of manuscripts (i.e., reviewers can now assess claims like “we examined the residuals and found that the assumptions of the model were met”) in many fields.

>>>>

---
title: Improving the communication and accessibility of stock assessment using interactive
  visualization tools
author: "Paul M. Regular, Gregory J. Robertson, Keith P. Lewis, etc?"
date: "Fisheries and Oceans Canada, Northwest Atlantic Fisheries Center, 80 East White Hills, St. John's, Newfoundland and Labrador, A1C 5X1, Canada"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
bibliography: references.bib
csl: apa.csl
---

Quantitative stock assessment plays a central role in modern fisheries management. Over time, and as new methods are developed, there is an increasing amount of data available to inform stock assessments. These data might be richer information on stock structure using an increasing array of markers and biomarkers, and/or the continued lengthening of population status and catch time series. Concurrently, analytical methods are now able to integrate many data sources into one stock assessment model. For well-monitored stocks, the challenge has shifted from having sufficient data and information to provide sound advice on stock status, to presenting the large quantity of information produced by increasingly complex statistical models to end-users in a meaningful way. Traditional methods of presenting information at stakeholder meetings and to fisheries managers are simply insufficient to convey the richness of the information available. More importantly, static presentation formats tend to stifle meaningful discussion of model results, and important questions and assumptions that should be discussed tend to get lost under the sheer volume of output. A solution to this problem of communicating complex output and results to encourage engaging discussion can be through using the effective use of interactive visualization tools. These are common tools we use every day on a range of web sites, but their use is no longer restricted to web site developers as these tools are being integrated into software commonly used by the research community.

Here we aim to demonstrate how interactive visualization tools provide an efficient and effective means of exploring and communicating the ever-expanding array of data inputs and model outputs. First, we focus on data that are commonly presented in stock assessments and use interactive maps to simplify the detailed exploration of data from a long-term tagging program. Second, we focus on the modeling aspect of stock assessment and demonstrate how dynamic and interactive data visualizations can be used to explore, diagnose and communicate results from an integrated assessment model. This structure corresponds to two important  steps in the stock assessment process: 1) knowing the data, and 2) data modeling.

@@

The data sets used in stock assessments are constantly growing due to the continuation of long-term monitoring efforts, the addition of new monitoring programs, or both. As such, stock assessment biologists need to manage large volumes of data from a variety of sources. Time series of reported landings and catch-at-age are often analyzed in conjunction with data from “fishery-independent” surveys that track changes in abundance and, in many cases, also monitor trends in biological factors such as age composition, growth rates, sex ratios and maturation stages. For some data-rich stocks, mark and recapture studies are also carried out to estimate movement, migration, growth rate, natural mortality, and discard mortality. All of the above-mentioned data sets are multidimensional and as the volume and variety of these data increases, it becomes more difficult to be aware of the details of each data set and discover key patterns within each. This challenge can be mediated, to a degree, by the application of interactive visualization tools as that they allow detailed exploration of the data behind a plot. For instance, the ability to zoom in on features or areas of interest and hover over specific points to reveal more information creates an interactive user-driven experience that expedites explorations of data. This is exemplified by an interactive mapping tool developed for the exploration of a long-term tagging study.

Northern cod (NAFO Divisions 2J3KL) has a rich history of tagging, starting in in 1954 [@taggart1995] and continuing to this day. The tagging and recovery data are captured in a standardized database, with fields typical of most tagging programs. This data base has over 600,000 records as of early 2019, with 2,000-10,000 tags deployed annually in recent decades. The tagging and capture data are used in the current assessment model for this stock [@cadigan2016], but tools to explore this extensive data set were limited, especially from a spatial perspective. To begin to explore and understand this large data set, we built a simple shiny [@chang2018a] application using the mapping package leaflet [@cheng2018], all within the framework of the R programming language [@R] and the RStudio IDE [@Rstudio]. As the application developed, it was incorporated into a shinydashboard [@chang2018b], as a means to quickly and dynamically subset the data (e.g. ranges of release and capture years, specific geographic locations). Given the large number of tags released, often at nearby sites, visualizing the data with static mapping was particularly challenging. The markercluster (https://github.com/Leaflet/Leaflet.markercluster) function available in leaflet was particularly useful as a means to dynamically scale the level of pooling of spatial points (Figure \@ref(fig:map)). This basic mapping tool allowed us to quickly become familiar with the data, identify outliers and incorrect data entries, as well as explore options on how to spatially pool the data for subsequent demographic analysis. Further tabs were added to provide basic summaries of the selected data (Figure \@ref(fig:summary)).

```{r map, echo = FALSE, fig.cap = '- Screen shot of cod tag mapping tool using shinydashboard and leaflet. The markercluster function dynamically splits or pools tagging locations (red, orange, green or yellow points) depending on zoom level, the recoveries positions (blue) are much fewer, and are left to be plotted individually at all scales. Options to include pop up labels are included, so specific information on each point can be retrieved with a mouse click (in this case: tag number, fish length, date released, and date captured), which is particularly useful when error checking.'}
knitr::include_graphics("figures/tagging_dashboard_map_tab.png")
```

@@

The Northern cod stock off southern Labrador and eastern Newfoundland is one of the most well studied stocks in eastern Canada, perhaps by virtue of its history. As such, there are multiple monitoring programs that help inform the status of the stock and data from most of these programs have been integrated into a state-space stock assessment model, called NCAM [@cadigan2016]. The model includes information from research vessel autumn trawl surveys (1983-present), Sentinel fishery surveys (1995-present), inshore acoustic surveys (1995-2009), fishery catch-at-age compositions and partial fishery landings (1983-present), and tagging (1983-present). Using a series of observation equations, this TMB based model reduces thousands of historical data points into quantities such as recruitment, spawning stock biomass, fishing mortality and natural mortality. Once the model is fit to the data, the next step is to produce visual representations of the data and model output. The usual approach would be to produce static presentations and documents with a series of figures and tables, but with large amounts of model inputs and outputs, this approach quickly becomes overwhelming for both the analyst and all the stakeholders involved. First, it is no longer feasible for the analyst to include and describe every figure and table produced in a single document. Second, it is difficult for stakeholders to efficiently digest the information that has been compressed into a series of static slides or pages. Interactive documents provide a potential solution to this problem as they allow much more information to be contained and accessible on a single screen.

In the pursuit of an easier and more efficient way to communicate results from NCAM, an interactive and self-contained “dashboard” was developed for the 2018 assessment of Northern cod [@dwyer2019]. The concept of using a dashboard was borrowed from the business community where dashboards are frequently used to group a series of interactive visuals and tables to provide at-a-glance views of key performance indicators. We used R-based packages [@R] and the RStudio IDE [@Rstudio] to construct a tool for exploring the input and output of NCAM, specifically the flexdashboard [@allaire2017] package to group interactive plotly-based [@sievert2018] visuals into a dynamic document. We also used the crosstalk [@cheng2016] package to link the data displayed across multiple plots. Via R Markdown [@allaire2018], the dashboard is rendered into a self-contained html file that is reproducible, interactive, and easy to update following modifications to the model or the addition of new data.

@@

```{r rv, echo = FALSE, fig.cap = '- Screenshot of the "RV survey" tab from the NCAM dashboard where total observed (dots) and NCAM model predicted values (lines) for the DFO RV survey index are shown in the upper left panel and scaled matrix plot of age-disaggregated standardized log residuals are shown in the lower left panel (blue = positive, red = negative, symbols scaled by size; grey = index values of zero). Age-disaggregated observed (dots) and predicted (lines) values from the DFO RV survey are shown in the right panel.'}
knitr::include_graphics("figures/NCAM_dashboard_rv-survey_tab.png")
```

@@

The amount of data available to scientists has grown by orders of magnitude in recent decades as has the complexities of data management, exploratory data analysis, in addition to the formal analyses and associated diagnostics [@lewis2018 and references therin]. The majority of this sequence of events, sometimes called “the data pipeline” [@leek2015a], have not traditionally been part of the peer-review process which sees only the end products of an analysis. However, decisions made along the data pipeline increasingly influence the outcome of the study. Gelman and Loken [-@gelman2014] coined the term “garden of forking paths” to illustrate that different conclusions can be arrived at depending on what decisions are made along during different stages of the analysis. Due to a number of limitations, such as available pages in journals, much of the data pipeline is not transparent nor is it reproducible. A number of authors have recently advocated for a culture of open science and reproducible research, i.e., a change in the transparency and reproducibility of science [@leek2015a; @leek2015b; @hampton2013; @hampton2015]. Proponents of open science and reproducible research highlight a number of benefits including a more productive and responsible scientific culture, an ability to address larger and more complex questions, as well as a more efficient workflow and ability to reproduce one’s own work [@lewis2018].

We  believe that the dashboard tool is a step forward in terms of open science and reproducible research in addition to being a major step forward in how stock assessments are presented and critiqued. While the entire data-pipeline can be revealed to all by sharing code and data, it is unlikely that an individual has the time to fully review and recreate a particular analysis in its entirety when most participants are already overcommitted [@banks2011]. In the case of NCAM, considerable statistical experience and expertise is required simply to run the model. However, the dashboard removes many of these difficulties. The html file contains all model inputs, outputs and diagnostics in an open and easy to use interactive document, and can be distributed prior to meetings and participants can assess the results and diagnostics at their leisure. Specific experience with the tools used to generate the results are not required to constructively and critically review of the results presented in the dashboard.

We must acknowledge there are costs to adopting the use of interactive tools. The first and obvious one being that staff are required to have both the time and training to effectively to develop these tools. In our two examples, the developers were relatively proficient with R and have natural aptitudes for programming, but had their formal training in field-based population ecology. Most programming issues were readily addressed through simple internet searchers, but some learning time was certainly needed (a few weeks). As more people in the fisheries community develop and use these tools, we expect tool develop speed and quality will only increase. It is our experience that the upfront investment in time has already paid dividends in terms of efficiencies of delivering products for subsequent stock assessments, and in the quality of the data exploration and understanding. Scientific exploration is rarely a linear process, and the ability to navigate around a dashboard fits the natural way people pursue ideas. In contrast, a slide presentation is effectively a linear and a single way of developing a scientific narrative; a method hardly conducive to exploring complex data problems.
